---
title: "Bivariate Viz"
---

Use this file for practice with the **bivariate viz** in-class activity.  Refer to the class website for details.
```{r}
library(tidyverse)
library(ggplot2)
```

*
**Exercise 2: 2 Quantitative Variables**
```{r}
# Load data
elections <- read.csv("https://mac-stat.github.io/data/election_2020_county.csv")

# Check it out
head(elections)
```


```{r}
# Set up the plotting frame
# How does this differ than the frame for our histogram of repub_pct_20 alone? x and y are defined by user
ggplot(elections, aes(y = repub_pct_20, x = repub_pct_16))
```
```{r}
# Add a layer of points for each county
# Take note of the geom!
ggplot(elections, aes(y = repub_pct_20, x = repub_pct_16)) +
  geom_point()
```


```{r}
# Change the shape of the points
# What happens if you change the shape to another number? Changes shape of the data points
ggplot(elections, aes(y = repub_pct_20, x = repub_pct_16)) +
  geom_point(shape = 3)
```

```{r}
# YOU TRY: Modify the code to make the points "orange"
# NOTE: Try to anticipate if "color" or "fill" will be useful here. Then try both.
ggplot(elections, aes(y = repub_pct_20, x = repub_pct_16)) +
  geom_point(color = "orange")
```
```{r}
# Add a layer that represents each county by the state it's in
# Take note of the geom and the info it needs to run!
ggplot(elections, aes(y = repub_pct_20, x = repub_pct_16)) +
  geom_text(aes(label = state_abbr))
```

**Exercise 4: Reflect**

-the strength of the relationship (weak/moderate/strong)
***Strong***

-the direction of the relationship (positive/negative)
***Positive***

-outliers (in what state do counties deviate from the national trend? Any ideas why this might be the case?) **
***Texas - some event that happened between 2016 and 2020***


**Exercise 4: Visualizing trend**

```{r}
ggplot(elections, aes(y = repub_pct_20, x = repub_pct_16)) +
  geom_point() +
  geom_smooth()
```
**Part a**

```{r}
ggplot(elections, aes(y = repub_pct_20, x = repub_pct_16)) +
  geom_smooth()
```
**Part b**

```{r}
ggplot(elections, aes(y = repub_pct_20, x = repub_pct_16)) +
  geom_point() +
  geom_smooth(method = "lm")
```
**Exercise 5: Your Turn**

```{r} 
ggplot(elections,(aes(y= repub_pct_20, x=median_rent))) +
  geom_point() +
  geom_smooth(method="lm")
```

```{r}
ggplot(elections, aes(y=repub_pct_20, x=median_age)) +
  geom_point()+
  geom_smooth(method="lm")
```
***-Since median_rent showed a stronger and clearer relationship (less noisy, more points seem to be close to the line of best fit) with repub_pct_20 than median_age, it is the better predictor.***

***-This makes sense because rent often reflects economic factors, urbanization, and cost of living, which are typically stronger indicators of voting patterns than median age.***



**Exercise 6: Sad Scatterplot**

```{r}
#repub_pct_20 is quantitative, but historical is categorical.
ggplot(elections, aes(y = repub_pct_20, x = historical)) +
  geom_point()
```
Explain why a scatterplot might not be an effective visualization for exploring this relationship

**A scatterplot is ineffective for visualizing relationships between a quantitative and a categorical variable because categorical data lacks a numeric axis (consists of groups rather than continuous numerical data), leading to overlapping points and poor pattern visibility.**


**Exercise 7: Quantitative vs Categorical – Violins & Boxes**

```{r}
# Side-by-side violin plots
ggplot(elections, aes(y = repub_pct_20, x = historical)) +
  geom_violin()
```


```{r}
# Side-by-side boxplots (defined below)
ggplot(elections, aes(y = repub_pct_20, x = historical)) +
  geom_boxplot()
```

 **Exercise 8: Quantitative vs Categorical – Intuition Check**
 
```{r}
ggplot(elections, aes(x = repub_pct_20)) +
  geom_density()
```
 More arguments and functions will be added in, in the exercise in order to stratify data by blue, purple and red categories.
 
 
 
 
 **Exercise 9: Quantitative vs Categorical – Density Plots**
```{r}
# Name two "bad" things about this plot
ggplot(elections, aes(x = repub_pct_20, fill = historical)) +
  geom_density()
```
*Two bad things about the plot:*

**- Color coding does not match the colors names of the different historical categories** 

**- Plots are overlapping each other**

```{r}
# What does scale_fill_manual do? Changes the colors of the plots manually
ggplot(elections, aes(x = repub_pct_20, fill = historical)) +
  geom_density() +
  scale_fill_manual(values = c("blue", "purple", "red"))
```
```{r}
# What does alpha = 0.5 do? Changes the opacity of the plots (made it more opaque)
# Play around with different values of alpha, between 0 and 1
ggplot(elections, aes(x = repub_pct_20, fill = historical)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("blue", "purple", "red"))
```
```{r}
# What does facet_wrap do?! Splits the data into subplots for clearer view
ggplot(elections, aes(x = repub_pct_20, fill = historical)) +
  geom_density() +
  scale_fill_manual(values = c("blue", "purple", "red")) +
  facet_wrap(~ historical)
```
```{r}
# Let's try a similar grouping strategy with a histogram instead of density plot.
# Why is this terrible? It doesnt really separate the 3 categories into subplots like the facet funtion
ggplot(elections, aes(x = repub_pct_20, fill = historical)) +
  geom_histogram(color = "white") +
  scale_fill_manual(values = c("blue", "purple", "red"))
```
 **Exercise 10**
 
*Out of boxplots, violins, and stratified density plots*
 
Which do you like best?

**Boxplots and stratified density plots**

What is one pro of density plots relative to boxplots?

**Density plots allow for a more nuanced visualization of the data’s distribution. They can show the full shape, including areas where the data is more concentrated or sparse, helping to reveal features like skewness or bimodality that boxplots cannot.**

What is one con of density plots relative to boxplots?

**Density plots can be sensitive to bandwidth choices, meaning that the appearance of the plot can change based on how the data is smoothed. This can lead to misleading or overly complex representations, while boxplots are simpler and provide clear information about central tendency and spread without this sensitivity.**


**Exercise 11: Categorical vs Categorical – Intuition Check**

```{r}
# Plot 1: adjust this to recreate the top plot
ggplot(elections, aes(x = historical, fill=winner_20)) +
  geom_bar()+
  scale_fill_manual(values = c("salmon", "turquoise"))
```
```{r}
ggplot(elections, aes(x=winner_20))+
  geom_bar()+
  facet_wrap(~historical)
```
**Exercise 12: Categorical vs Categorical**


```{r}
# A stacked bar plot
# How are the "historical" and "winner_20" variables mapped to the plot, i.e. what roles do they play?
ggplot(elections, aes(x = historical, fill = winner_20)) +
  geom_bar()
```
```{r}
# A faceted bar plot
ggplot(elections, aes(x = winner_20)) +
  geom_bar() +
  facet_wrap(~ historical)
```

```{r}
# A side-by-side bar plot
# Note the new argument to geom_bar
ggplot(elections, aes(x = historical, fill = winner_20)) +
  geom_bar(position = "dodge")
```

```{r}
# A proportional bar plot
# Note the new argument to geom_bar
ggplot(elections, aes(x = historical, fill = winner_20)) +
  geom_bar(position = "fill")
```


Part a
Name one pro and one con of using the “proportional bar plot” instead of one of the other three options.

**Pro of using a proportional bar plot:**
Proportional bar plots are great for showing the relative sizes of categories within a whole, making it easier to compare the proportions of different categories across groups. This is especially useful when you want to emphasize the distribution of categories in relation to the total.

**Con of using a proportional bar plot:**
Proportional bar plots can make it harder to compare absolute values, as they focus on percentages or proportions rather than actual counts. If you're interested in the precise number of occurrences in each category, other types of bar plots (like side-by-side or stacked) might be more informative.

Part b
What’s your favorite bar plot from part and why?

**Faceted Bar Plot - because it is easied to look at the visualization when data is stratified as much as possible. It does not look crowded, overlapped or noisy.**

**Exercise 14: Practice

```{r}
weather <- read.csv("https://mac-stat.github.io/data/weather_3_locations.csv")
```

```{r}
head(weather)
```


```{r}
# How do 3pm temperatures (temp3pm) differ by location?
ggplot (weather, aes(x=temp3pm, fill = location)) +
  geom_density()+
  facet_wrap(~location)
  
```

```{r}
# How might we predict the 3pm temperature (temp3pm) by the 9am temperature (temp9am)?
ggplot(weather, aes(x=temp9am, y=temp3pm))+
  geom_point()+
  geom_smooth(method="lm")
```

```{r}
# How do the number of rainy days (raintoday) differ by location?
ggplot(weather, aes(x=raintoday, fill=location))+
  geom_bar()+
  facet_wrap(~location)
  
```

